NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
Edits to the 2 first sections of results

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
electrostatic interactions and direct binding, and can induce
membrane fusion and phase separation in the presence of calcium ions.
Molecular details of these phenomena are not well understood,
because accurate models to interpret the experimental data have not
been available. Here, we use solid-state NMR S-DROSS to measure the signs of the experimental POPS headgroup
order parameters, and gather a set of experimental NMR results from pure PS and mixed PC:PS lipid bilayers. Using the open collaboration
method, we then extract data from a wide range of available PS MD models (force fields) and asses the PS headgroup structures and the ion binding produced by these force fields by comparing to the NMR data. We find that none of the models reproduce the NMR data within experimental accuracy,
but the best ones suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity---the response of PS headgroups to bound ions can even
\emph{qualitatively} differ from experiments. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    The data for DOPS is at 303~K and the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36 --- the force field producing the best agreement with experiments ---
     demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups. The publicly available trajectory was used for CHARMM36 POPC~\cite{POPCcharmm36T303K}.
  }
%  \todo{Figure clarity is questioned in: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/48}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
    Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

The different PS MD models produce a wide variety of headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
as previously observed also for PC lipids \cite{botan15}. None of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups.

@@

However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and narrow angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.

>>>>

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with signaling and other proteins via
electrostatic interactions and direct binding, and can induce
membrane fusion and phase separation in the presence of calcium ions.
Molecular details of these phenomena are not well understood,
because accurate models to interpret the experimental data have not
been available. Here, we use solid-state NMR S-DROSS experiments to measure the signs of the previous set of experimental NMR
order parameter magnitudes available for POPS to assess the phosphoserine structure in the different atomistic force-fields presently used.    
We also gather a set of experimental NMR data that can be used to assess the ion binding to pure PS and mixed PC:PS lipid bilayers
in molecular dynamics (MD) simulations.
Using the open collaboration
method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
%or: Using the open collaboration method, we collect a database of MD simulations using wide range of available PS models (force fields) and perform a comparison to NMR results
We find that none of the models reproduce the NMR data within experimental accuracy,
but the best ones suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity.
Furthermore, the response of PS headgroups to bound ions
\emph{qualitatively} differs from experiments. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13} whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are without additional salt.
    The data for DOPS is at 303~K and the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations around
    (A) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and (B) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at~\citenum{POPCcharmm36T303K}.
  }

\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough, subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
    Here M indicates a magnitude problem, F a forking problem; letter size increases with problem severity. Color scheme: within experimental error (dark green), almost within experimental errorÓ (light green), Òclear deviation from experiments (light red), and major deviation from experiments (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see SI section~\ref{Ranking}.
  }
\end{figure}

The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups. However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

@@

Interestingly, the three models that best fit the experimental data have a narrower distribution in the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral angle (single peak around 120$^{\circ}$) in comparison to the other models which yield a distribution between two angles (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and narrow angle distribution for PS than for PC headgroup in
CHARMM36 simulations that have the best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
% Also, the sampled conformations of glycerol backbone significantly vary between different simulation models (Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),but further analysis is beyond the scope of this work which focuses on the PS headgroup. There is very little actual information in this statement. The variance was already noted above

These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.
